Red, Blue, or Purple? 2024’s Battleground States Tell a Story

### 2024-11-04 paper/paper.qmd
paper formatting finished

---
title: "Red, Blue, or Purple? 2024’s Battleground States Tell a Story"
subtitle: "Pennsylvania, Ohio, and Nevada Become Game Changers as Trump Gains Ground"
author: 
  - Yingke He
  - Ziheng Zhong
thanks: "Code and data are available at: [https://github.com/iJustinn/Election_Prediction](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "This paper analyzes polling data for the 2024 U.S. Presidential election, examining key candidates, notably Donald Trump and Kamala Harris, to gain a comprehensive perspective on the race. A Hierarchical Bayesian Model is applied in this study, and the findings reveal significant variations across states, with Pennsylvania, Ohio, Georgia, and Nevada emerging as particularly influential battlegrounds that ultimately tilt in favor of Trump. By highlighting these trends, this research underscores the dynamic and evolving nature of the U.S. political landscape and the crucial role that swing states play in determining election outcomes."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

```{r}
#| label: fig-pollscore_distribution
#| fig-cap: "Distribution of Pollscore" 
#| echo: false
#| warning: false
#| message: false

@@

# Plot histogram with overlayed density plot
ggplot(data, aes(x = pollscore)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.05, fill = "steelblue", color = "white", alpha = 0.7) +
  geom_density(fill = "skyblue", alpha = 0.5) +
  labs(x = "Poll Score",
       y = "Density") +
  theme_minimal()

@@

Let: 
\begin{align*}
y_{ijk} & : \text{ The target variable, representing the actual election vote share for candidate } k \text{ in state } \\
& \quad i \text{ with pollster } j. \\ 
\text{pct}_{ijk} & : \text{ The observed polling percentage for candidate } k \text{ in state } i \text{ by pollster } j. \\ 
\text{sample\_size}_{ijk} & : \text{ The sample size of the poll, which helps in weighing the poll reliability.} \\ 
\text{days\_to\_election}_{ijk} & : \text{ Derived as the days remaining until the election from the poll’s end date, capturing } \\
& \quad \text { the effect of recency.} \\ 
\text{pollscore}_{ijk} & : \text{ The reliability score for each pollster } j, \text{ reflecting the accuracy of the pollster's } \\
& \quad \text { historical results.} \\
\end{align*}

The model takes the form of the following equation:
\begin{align}
y_{ijk} = \alpha_i + \beta_j + \gamma_k + \delta \cdot \text{pct}_{ijk} + \eta \cdot \text{sample\_size}_{ijk} + \theta \cdot \text{days\_to\_election}_{ijk} + \zeta \cdot \text{pollscore}_{ijk} + \epsilon_{ijk} 
\end{align}

where:
\begin{align*}
\alpha_i &\sim \mathcal{N}(\mu_{\alpha}, \sigma_{\alpha}^2): \quad \text{State-level random effect for each state } i, \text{ capturing regional variations in voting patterns.} \\ 
\beta_j &\sim \mathcal{N}(\mu_{\beta}, \sigma_{\beta}^2): \quad \text{Pollster-level random effect for each pollster } j, \text{ accounting for systematic biases or } \\
& \quad \text {differences in polling methods.} \\ 
\epsilon_{ijk} &\sim \mathcal{N}(0, \sigma^2): \quad \text{Error term, accounting for random noise.} \\ 
\gamma_k &: \quad \text{Candidate fixed effect for each candidate } k, \text{ representing baseline support across states and pollsters.} \\ 
\delta &: \quad \text{Coefficient for Percentage of Votes by Poll } (\text{pct}_{ijk}), \text{ reflecting how poll support translates to actual } \\
& \quad \text {vote share.} \\ 
\eta &: \quad \text{Coefficient for Sample Size } (\text{sample\_size}_{ijk}), \text{ weighing polls based on their reliability.} \\ 
\theta &: \quad \text{Coefficient for Days to Election } (\text{days\_to\_election}_{ijk}), \text{ capturing the trend in support as the election } \\
& \quad \text { date approaches.} \\ 
\zeta &: \quad \text{Coefficient for Pollscore } (\text{pollscore}_{ijk}), \text{ accounting for the reliability of the pollster's } \\
& \quad \text { historical performance.} \\
\end{align*}

@@

\begin{align*}
\alpha_i &: \quad \text{Captures state-specific effects, adjusting the baseline vote share prediction based on regional differences.} \\ 
\beta_j &: \quad \text{Accounts for systematic biases or differences in methodologies across pollsters.} \\ 
\gamma_k &: \quad \text{Provides an overall baseline effect for each candidate, independent of state or pollster.} \\ 
\delta &: \quad \text{Measures how closely polling support translates to actual vote share.} \\ 
\eta &: \quad \text{Adjusts the model’s sensitivity to polls based on sample size, giving more weight to larger polls.} \\ 
\theta &: \quad \text{Captures how support trends change as the election date approaches.} \\ 
\zeta &: \quad \text{Incorporates the reliability of the pollster based on historical performance, weighing polls by } \\
& \quad \text { the pollster's accuracy.} \\ 
\end{align*}

@@

\begin{align*}
\alpha_i &\sim \mathcal{N}(0, 2.5): \quad \text{State-level random effect prior for each state } i. \\ 
\beta_j &\sim \mathcal{N}(0, 2.5): \quad \text{Pollster-level random effect prior for each pollster } j. \\ 
\gamma_k &\sim \mathcal{N}(0, 2.5): \quad \text{Candidate fixed effect prior for each candidate } k. \\ 
\delta, \eta, \theta, \zeta &\sim \mathcal{N}(0, 1): \quad \text{Coefficients for polling percentage, sample size, days to election, and pollscore.} \\ 
\sigma &\sim \text{Exponential}(1): \quad \text{Prior for the standard deviation of the error term.} \\
\end{align*}

@@

```
Shown in above @tbl-candidate_fixed_effect, the candidate fixed effects in this model capture the baseline level of support for each candidate, reflecting their overall popularity or favorability across the polls. For example, Kamala Harris has a candidate fixed effect of 47.22, while Donald Trump has a slightly lower effect of 45.28. These values suggest that, on average, Harris may have a marginally higher baseline level of support in the polling data compared to Trump. These fixed effects allow the model to adjust for inherent differences in candidate popularity, providing a more accurate prediction by factoring in each candidate's overall favorability independent of state, pollster, or specific poll characteristics. This adjustment is crucial for capturing the candidates' general standing in the electorate, contributing to more reliable forecasting.

@@

The final predictions for the 2024 presidential election indicate a close race between the two candidates. The model results are summarized in @tbl-modelresults.Based on the weighted average calculations, which incorporate factors such as state effects, pollster reliability, candidate-specific adjustments, and recent polling trends, Donald Trump is predicted to receive approximately 46.56% of the vote, while Kamala Harris is predicted to receive around 45.65%. These predictions account for adjustments based on sample size, recency of the polls, and reliability scores, ensuring that more recent and reliable polls have a greater influence on the final outcome. The model also applies a slight adjustment to account for known biases, providing a robust estimation of each candidate’s standing. Although these predictions offer a snapshot of the current polling landscape, they should be interpreted with caution, as last-minute changes in voter sentiment could still impact the final outcome.

@@

```{r}
#| label: tbl-modelresults_states
#| echo: false
#| warning: false
#| message: false

@@

# Display the results in a table
summary_data %>%
  kable(
    col.names = c("State", "Trump Predicted (%)", "Harris Predicted (%)", "Predicted Winner"),
    caption = "Model Prediction Results for Key States"
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down", "hold_position"))

@@

```{r}
#| label: fig-modelresults_states
#| echo: false
#| warning: false
#| message: false

@@

# Plotting
ggplot(swing_states_long, aes(x = State, y = Predicted_Percentage, fill = Candidate)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_manual(values = c("Trump_Predicted" = "red", "Harris_Predicted" = "blue"),
                    labels = c("Trump Predicted (%)", "Harris Predicted (%)")) +
  labs(
    title = "Model Prediction Results for Key Swing States",
    x = "States with Close Predicted Vote Share",
    y = "Predicted Percentage (%)",
    fill = "Candidate"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

```{r}
#| label: fig-modelresults_map
#| echo: false
#| warning: false
#| message: false
#| fig-width: 20
#| fig-height: 8
#| fig.fullwidth: true

@@

# Plot using usmap with updated trends, specific legend labels, and state names
plot_usmap(data = state_wide, values = "color", labels = TRUE, label_color = "white") +
  scale_fill_manual(
    values = c("red" = "red", "blue" = "blue", "gray" = "gray"), 
    labels = c("red" = "Trump", "blue" = "Harris", "gray" = "Swing States")  # Set labels correctly
  ) +
  labs(
    title = "2024 U.S. Presidential Election Prediction by State",
    subtitle = "Predicted Winner by State with Key Swing States Highlighted",
    fill = "Predicted Winner"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

@@

The updated map @fig-modelresults_map above provides a visual forecast of the 2024 U.S. Presidential Election by state, using weighted polling data and recent trend adjustments. Each state is color-coded to show the projected winner: red for states likely to support Trump, blue for those leaning towards Harris, and gray for swing states where the predicted vote margins are extremely close. This map shows a competitive electoral landscape. Harris is projected to perform well in traditionally Democratic regions on the West Coast and in the Northeast, including states like California, New York, and Illinois. Trump demonstrates strong support across the South, Midwest, and several central states, such as Texas, Florida, and the Dakotas.

@@

#| label: fig-state_effect
#| fig-cap: "State Effect" 
#| echo: false
#| warning: false
#| message: false

# Load the state effect data
state_effect <- read_csv(here("data", "02-analysis_data", "state_effect.csv"))

# Check the data structure
glimpse(state_effect)

@@

```{r}
#| label: fig-pollster_effect
#| fig-cap: "Relationship Between Poll Score Reliability and Support Percentage for Kamala Harris" 
#| echo: false
#| warning: false
#| message: false

@@

# 1. Histogram of Pollster Effects
ggplot(pollster_effects, aes(x = pollster_effect)) +
  geom_histogram(bins = 10, fill = "steelblue", color = "black") +
  labs(
    title = "Distribution of Pollster Effects",
    x = "Pollster Effect",
    y = "Number of Pollsters"
  ) +
  theme_minimal()
```

@@

```{r}
#| label: fig-pollscore_pct_harris
#| fig-cap: "Relationship Between Poll Score Reliability and Support Percentage for Kamala Harris" 
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2
#| layout-nrow: 1

>>>>

---
title: "Red, Blue, or Purple? 2024’s Battleground States Tell a Story"
subtitle: "Pennsylvania, Ohio, and Nevada Become Game Changers as Trump Gains Ground"
author: 
  - Yingke He
  - Ziheng Zhong
thanks: "Code and data are available at: [https://github.com/iJustinn/Election_Prediction.git](https://github.com/iJustinn/Election_Prediction.git)."
date: today
date-format: long
abstract: "This paper analyzes polling data for the 2024 U.S. Presidential election, examining key candidates, notably Donald Trump and Kamala Harris, to gain a comprehensive perspective on the race. A Hierarchical Bayesian Model is applied in this study, and the findings reveal significant variations across states, with Pennsylvania, Ohio, Georgia, and Nevada emerging as particularly influential battlegrounds that ultimately tilt in favor of Trump. By highlighting these trends, this research underscores the dynamic and evolving nature of the U.S. political landscape and the crucial role that swing states play in determining election outcomes."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

```{r}
#| label: fig-pollscore_distribution
#| fig-cap: "Distribution of Pollscore" 
#| echo: false
#| warning: false
#| message: false
#| fig-height: 3
#| fig-width: 5

@@

# Plot histogram with overlayed density plot
ggplot(data, aes(x = pollscore)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.05, fill = "darkgrey", color = "white", alpha = 0.7) +
  geom_density(fill = "lightgrey", alpha = 0.5) +
  labs(x = "Poll Score",
       y = "Density") +
  theme_minimal()

@@

Let: 
\begin{align*}
y_{ijk} & : \text{ The target variable, representing the actual election vote share for} \\
& \quad \text {candidate } k \text{ in state } i \text{ with pollster } j. \\ 
\text{pct}_{ijk} & : \text{ The observed polling percentage for candidate } k \text{ in state } i \text{ by pollster } j. \\ 
\text{sample\_size}_{ijk} & : \text{ The sample size of the poll, which helps in weighing the poll reliability.} \\ 
\text{days\_to\_election}_{ijk} & : \text{ Derived as the days remaining until the election from the poll’s end date, } \\
& \quad \text {capturing the effect of recency.} \\ 
\text{pollscore}_{ijk} & : \text{ The reliability score for each pollster } j, \text{ reflecting the accuracy of the } \\
& \quad \text { pollster's historical results.} \\
\end{align*}

The model takes the form of the following equation:
\begin{align}
y_{ijk} = \alpha_i + \beta_j + \gamma_k + \delta \cdot \text{pct}_{ijk} + \eta \cdot \text{sample\_size}_{ijk} +  \\ 
\theta \cdot \text{days\_to\_election}_{ijk} + \zeta \cdot \text{pollscore}_{ijk} + \epsilon_{ijk} 
\end{align}

where:
\begin{align*}
\alpha_i &\sim \mathcal{N}(\mu_{\alpha}, \sigma_{\alpha}^2): \quad \text{State-level random effect for each state } i, \text{ capturing regional variations in } \\ 
& \quad \text{voting patterns.} \\ 
\beta_j &\sim \mathcal{N}(\mu_{\beta}, \sigma_{\beta}^2): \quad \text{Pollster-level random effect for each pollster } j, \text{ accounting for systematic } \\
& \quad \text {biases or differences in polling methods.} \\ 
\epsilon_{ijk} &\sim \mathcal{N}(0, \sigma^2): \quad \text{Error term, accounting for random noise.} \\ 
\gamma_k &: \quad \text{Candidate fixed effect for each candidate } k, \text{ representing baseline support across states} \\ 
& \quad \text{ and pollsters.} \\
\delta &: \quad \text{Coefficient for Percentage of Votes by Poll } (\text{pct}_{ijk}), \text{ reflecting how poll support translates } \\
& \quad \text {to actual vote share.} \\ 
\eta &: \quad \text{Coefficient for Sample Size } (\text{sample\_size}_{ijk}), \text{ weighing polls based on their reliability.} \\ 
\theta &: \quad \text{Coefficient for Days to Election } (\text{days\_to\_election}_{ijk}), \text{ capturing the trend in support } \\
& \quad \text { as the election date approaches.} \\ 
\zeta &: \quad \text{Coefficient for Pollscore } (\text{pollscore}_{ijk}), \text{ accounting for the reliability of the pollster's } \\
& \quad \text { historical performance.} \\
\end{align*}

@@

\begin{align*}
\alpha_i &: \quad \text{Captures state-specific effects, adjusting the baseline vote share prediction based on } \\ 
& \quad \text{ regional differences.} \\
\beta_j &: \quad \text{Accounts for systematic biases or differences in methodologies across pollsters.} \\ 
\gamma_k &: \quad \text{Provides an overall baseline effect for each candidate, independent of state or pollster.} \\ 
\delta &: \quad \text{Measures how closely polling support translates to actual vote share.} \\ 
\eta &: \quad \text{Adjusts the model’s sensitivity to polls based on sample size, giving more weight to } \\ 
& \quad \text{  larger polls.} \\
\theta &: \quad \text{Captures how support trends change as the election date approaches.} \\ 
\zeta &: \quad \text{Incorporates the reliability of the pollster based on historical performance, weighing } \\
& \quad \text { polls by the pollster's accuracy.} \\ 
\end{align*}

@@

\begin{align*}
\alpha_i &\sim \mathcal{N}(0, 2.5): \quad \text{State-level random effect prior for each state } i. \\ 
\beta_j &\sim \mathcal{N}(0, 2.5): \quad \text{Pollster-level random effect prior for each pollster } j. \\ 
\gamma_k &\sim \mathcal{N}(0, 2.5): \quad \text{Candidate fixed effect prior for each candidate } k. \\ 
\delta, \eta, \theta, \zeta &\sim \mathcal{N}(0, 1): \quad \text{Coefficients for polling percentage, sample size, days to } \\ 
& \quad \text{ election, and pollscore.} \\
\sigma &\sim \text{Exponential}(1): \quad \text{Prior for the standard deviation of the error term.} \\
\end{align*}

@@

```
Shown in @tbl-candidate_fixed_effect, the candidate fixed effects in this model capture the baseline level of support for each candidate, reflecting their overall popularity or favorability across the polls. For example, Kamala Harris has a candidate fixed effect of 47.22, while Donald Trump has a slightly lower effect of 45.28. These values suggest that, on average, Harris may have a marginally higher baseline level of support in the polling data compared to Trump. These fixed effects allow the model to adjust for inherent differences in candidate popularity, providing a more accurate prediction by factoring in each candidate's overall favorability independent of state, pollster, or specific poll characteristics. This adjustment is crucial for capturing the candidates' general standing in the electorate, contributing to more reliable forecasting.

@@

The final predictions for the 2024 presidential election indicate a close race between the two candidates. The model results are summarized in @tbl-modelresults. Based on the weighted average calculations, which incorporate factors such as state effects, pollster reliability, candidate-specific adjustments, and recent polling trends, Donald Trump is predicted to receive approximately 46.56% of the vote, while Kamala Harris is predicted to receive around 45.65%. These predictions account for adjustments based on sample size, recency of the polls, and reliability scores, ensuring that more recent and reliable polls have a greater influence on the final outcome. The model also applies a slight adjustment to account for known biases, providing a robust estimation of each candidate’s standing. Although these predictions offer a snapshot of the current polling landscape, they should be interpreted with caution, as last-minute changes in voter sentiment could still impact the final outcome.

@@

```{r}
#| label: tbl-modelresults_states
#| tbl-cap: "Model Prediction Results for Key States"
#| echo: false
#| warning: false
#| message: false

@@

# Display the results in a table
summary_data %>%
  kable(
    col.names = c("State", "Trump Predicted (%)", "Harris Predicted (%)", "Predicted Winner"),
  ) %>%
  kable_styling(latex_options = c("striped"))

@@

```{r}
#| label: fig-modelresults_states
#| fig-cap: "Model Prediction Results for Key Swing States"
#| echo: false
#| warning: false
#| message: false

@@

# Plotting
ggplot(swing_states_long, aes(x = State, y = Predicted_Percentage, fill = Candidate)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_manual(values = c("Trump_Predicted" = "red", "Harris_Predicted" = "blue"),
                    labels = c("Harris Predicted (%)", "Trump Predicted (%)")) +
  labs(
    x = "States with Close Predicted Vote Share",
    y = "Predicted Percentage (%)",
    fill = "Candidate"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

```{r}
#| label: fig-modelresults_map
#| fig-cap: "2024 U.S. Presidential Election Prediction by State"
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 7

@@

# Plot using usmap with updated trends, specific legend labels, and state names
plot_usmap(data = state_wide, values = "color", labels = TRUE, label_color = "white") +
  scale_fill_manual(
    values = c("red" = "red", "blue" = "blue", "gray" = "gray"), 
    labels = c("red" = "Trump", "blue" = "Harris", "gray" = "Swing States")  # Set labels correctly
  ) +
  labs(
    fill = "Predicted Winner",
    caption = "Predicted Winner by State with Key Swing States Highlighted"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

@@

The updated map @fig-modelresults_map provides a visual forecast of the 2024 U.S. Presidential Election by state, using weighted polling data and recent trend adjustments. Each state is color-coded to show the projected winner: red for states likely to support Trump, blue for those leaning towards Harris, and gray for swing states where the predicted vote margins are extremely close. This map shows a competitive electoral landscape. Harris is projected to perform well in traditionally Democratic regions on the West Coast and in the Northeast, including states like California, New York, and Illinois. Trump demonstrates strong support across the South, Midwest, and several central states, such as Texas, Florida, and the Dakotas.

@@

```{r}
#| label: fig-state_effect
#| fig-cap: "State Effect" 
#| echo: false
#| warning: false
#| message: false

# Load the state effect data
state_effect <- read_csv(here("data", "02-analysis_data", "state_effect.csv"), show_col_types = FALSE)

@@

```{r}
#| label: fig-pollster_effect
#| fig-cap: "Distribution of Pollster Effects" 
#| echo: false
#| warning: false
#| message: false
#| fig-height: 3
#| fig-width: 5

@@

# 1. Histogram of Pollster Effects
ggplot(pollster_effects, aes(x = pollster_effect)) +
  geom_histogram(bins = 10, fill = "darkgrey", color = "black") +
  labs(
    x = "Pollster Effect",
    y = "Number of Pollsters"
  ) +
  theme_minimal()
```

@@

```{r}
#| label: fig-pollscore_pct_harris
#| fig-cap: "Relationship Between Poll Score Reliability and Support Percentage for Kamala Harris" 
#| echo: false
#| warning: false
#| message: false
#| fig-height: 4
#| fig-width: 4

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2
#| layout-nrow: 1
#| fig-height: 3
#| fig-width: 4
